---
title: "Scatter Plot"
author: "Jaxson Freund"
format: html
editor: source
---

## Context
From January 1-15, 2024, I visited Costa Rica as a class trip. I conducted an observational study on foraging behavior of leafcutter ants.  My goal is to assess qualities of leaves that are rejected far from the colony, vs. leaves that are carried all the way to the colony entrance and rejected at the doorstep.

## Data
```{r}
# vector x (distance from main colony entrance)
x <- c(0, 6, 12, 18, 3, 9, 15, 21, 1, 7, 13, 19, -2, 4, 10, 16, 22)

# vector y (leaf pass in quadrat)
y <- c(4, 1, 0.5, 0.2, 2.5, 0.5, 0.3, 0.2, 3.9, 0.6, 2.1, 0.3, 0.7, 1.2, 1.4, 0.9, 0.4)

# vector z (leaf count in quadrat)
z <- c(124, 43, 21,  9, 118, 22, 9, 7, 137, 46, 95, 15, 36, 76, 88, 44,  21)

```

## Sort based on X
```{r}
#create a data  frame
data <- data.frame(x = x, y = y, z = z)

# Use order to get the indices that would sort the data frame by x
order_indices <- order(data$x)

# Use the indices to reorder the data frame
sorted_data <- data[order_indices, ]
```
## Display the sorted data
```{R}
# Display the sorted data frame
print(sorted_data)
```
## Format the table
```{R}
#change row headers
colnames(sorted_data) <-
c("Distance (m)", "Leaf Mass (g)", "Leaf Count")
#remove indices on the left side
print(sorted_data, row.names = FALSE)

```

## Plot Leaf Count vs. Distance
```{R}
plot(x, y, main = "Leaf Mass vs. Distance from Main Colony Entrance", xlab = "Distance (m)", ylab = "Leaf Mass (g)", col = "blue", pch = 16)

#Add an exponential line of best fit
fit <- lm(log(y) ~ x)

# Generate predicted values for the exponential fit
x_pred <- seq(min(x), max(x), length.out = 100)
y_pred <- exp(predict(fit, newdata = data.frame(x = x_pred)))

# Add the exponential line to the plot
lines(x_pred, y_pred, col = "blue", type = "l")

#add the equation of best fit
# Extract coefficients and transform back from log scale
a <- exp(coef(fit)[1])
b <- coef(fit)[2]

# Print the equation on the plot
eq <- sprintf("y = %.2f * exp(%.2fx)", a, b)
text(3, max(y), eq, pos = 4, col = "black", cex = 0.8)
```
## Plot Leaf Count vs. Distance
```{R}
plot(x, z, main = "Leaf Count vs. Distance from Main Colony Entrance", xlab = "Distance (m)", ylab = "Leaf Count", col = "red", pch = 16)
fit <- lm(log(z) ~ x)
x_pred <- seq(min(x), max(x), length.out = 100)
z_pred <- exp(predict(fit, newdata = data.frame(x = x_pred)))
lines(x_pred, z_pred, col = "red", type = "l")
a <- exp(coef(fit)[1])
b <- coef(fit)[2]
eq <- sprintf("z = %.2f * exp(%.2fx)", a, b)
text(3, max(z), eq, pos = 4, col = "black", cex = 0.8)
```